Mergeable Types for Highly-Available Replicated Data

### 2017-02-23 paper/semantics.tex
Semantics.

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_1$ at the instance of
merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
($v_{22}$) could be a common ancestor for a version on $b_1$
($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
($v_{32}$) could be a common ancestor for the new version on $b_1$
($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
ancestor of $v_{13}$, and both ancestors are not ordered by the
ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
$v_{44}$. We observe that this scenario can be prevented if, when
merging $b_3$, $b_1$ insists on an ancestor relation between the
merged version ($v_{22}$) of the $b_2$, the previously merged branch,
and the latest version of $b_3$, the current merging branch. We call
the last merged version the \emph{external locus} of the branch. By
requiring that, for every branch $b$, the external locii of $b$ at
various points in time (i.e., external locus of every prefix of $b$)
be totally ordered, we effectively enforce the total order among
versions that have more than one common ancestor, thereby preventing
the possibility of multiple LCAs.

%% Multiple versions that merge any pair of versions are totally
%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.

@@

\begin{definition} [\bfseries Least Internal and External Ancestors]
Given a branch $b$ and a version $v\in b$, least internal ancestor is
the immediate ancestor version ($v_i$) of $v$ on $b$. Least external
ancestor ($v_o$) is an external ancestor of $v$ that is not an
ancestor of any other external ancestor of $v$. That is,
$\under{H}{v_o \preceq v}$, and there does not exist a $v_o' \not\in
b$ such that $\under{H}{v_o' \preceq v}$ and $\under{H}{v_o \preceq
v_o'}$. 
% We lift the definition to the level of branches; $v_o$ is an
% external locus of a branch $b$ iff it is an external locus of its
% latest version.
\end{definition}

Note that all branches (except $H(t_{\top})$) begin with a \C{FORK}
tag, hence every version has at least one least external ancestor. We
later prove that in the histories generated by the rules in
Fig.~\ref{fig:opsem}, the external ancestors of every version are
totally ordered, hence there is exactly one least external ancestor
for each version. We call the unique least external ancestor of a
version as its \emph{external locus}. Least internal ancestor, by
definition, is unique, which we sometimes call \emph{internal locus}.

@@

\begin{lemma} [\bfseries Unique External Locus]
Every version $v$ (except the \C{INIT} version) in the history $H$
generated by the rules in Fig.~\ref{fig:opsem} has a unique external
locus.
\end{lemma}
\begin{proof}
By case analysis on the reduction relation. In each case, we assume
that the unique external locus property holds for the initial history,
and prove that it holds for the final history.

In \rulelabel{E-Run} case, initial version is the unique external locus of
the new version on the new branch $H(t)$. In \rulelabel{E-Fork} case,
which creates the branch $H(t')$ by forking from $H(t)$, the latest
version on $H(t)$ is the external locus of the first version on
$H(t)$. In \rulelabel{E-Push} case, the version on $H(t)$ has the same
external locus as the previous latest version, which, premise says, is
unique. No new version is created in \rulelabel{E-Pull}. In
\rulelabel{E-Pull-Wait}, new version is created by merging the latest
version ($v'$) on $H(t')$ into the latest version $v$ on $H(t)$.
Premise says that $v$ has a unique external locus (call it $v_o$). The
new version $v_m$ inherits $v_o$ as a least external ancestor, and
also has $v'$ as another least external ancestor. However, since
$\under{H}{v' \mbleto v}$, we have $\under{H}{v_o \preceq v'}$, we
have $v'$ as the the external locus.
\qed
\end{proof}

%% If we are allowed to merge earlier versions on branches, then we
%% have a possibility of utmost two LCAs.

\begin{theorem} [\bfseries Unique LCA]
Every pair of versions $v_1$ and $v_2$ in the history $H$ generated by
the rules in Fig.~\ref{fig:opsem} have a unique least
common ancestor. 
\end{theorem}
\begin{proof}
By case analysis on the step relation. In each case, we assume that
the unique LCA property holds for the initial history, and prove that
it holds for the final history.

The \rulelabel{E-Run} case is trivial. The \rulelabel{E-Push} case
creates a new version ($v$) with \C{PUSH} tag at the end of a branch.
Let $v_1$ be the previous version on the branch, and let $v_2$ be any
other version. The premise is that $v_1$ and $v_2$ have a unique LCA
$v_0$. Observe that $v$'s internal locus is $v_1$ and its external
locus is same as that of $v_1$. Hence, all ancestors of $v_1$
(including itself) are also ancestors of $v$. Moreover, $v$ has no new
ancestors except itself, which is the latest version being added.
Hence, any common ancestor of $v_1$ and $v_2$ is also a common
ancestor of $v$ and $v_2$, while there cannot be new common ancestors.
Hence LCA of $v$ and $v_2$ is unique. The reasoning for
\rulelabel{E-Fork} is similar. The \rulelabel{E-Pull} case is trivial.

The \rulelabel{E-Pull-Wait} case is where the merge happens. Version
$v'$ is merging into version $v$ to produce the next version $v_m$.
Let $v''$ be any version. From the premise, we know that $v$ and $v''$
have a unique LCA (call it $v_1$), and $v'$ and $v''$ have a unique
LCA (call it $v_2$). Also, we know that $v''$ has a unique external
locus (call it $v_3$). We now consider cases
\begin{itemize}
  \item $v_1$ and $v_2$ are external to both $v$ and $v'$: Hence $v_1$
  is an external ancestor of $v$, and $v_2$ is an external ancestor
  of $v'$. Since $v$ and $v'$ are mergeable, there must exist a $v_4$
  such that $\under{H}{v_1 \preceq v_4}$ and $\under{H}{v_1 \preceq
  v_4}$, and either $\under{H}{v_4 \preceq v}$ or $\under{H}{v_4
  \preceq v'}$.  are external ancestors of
  $v_2$. Hence,
  $\under{H}{v_0 \preceq v_3} \conj \under{H}{v_1 \preceq v_3}$.
  Since $\under{H}{v_0 \preceq v_m} \conj \under{H}{v_1 \preceq v_m}$, 
  $v_2$ and $v_m$ have a unique LCA.

  \item $v_0$ is an internal ancestor of $v_2$, hence an external
  ancestor of $v$, hence an external ancestor of $v_m$. $v_1$ is
  an ancestor of $v'$, hence also an external ancestor of $v_m$. 
\end{itemize}

As proved
before, the merged version ($v_m$) has a unique external locus, which
is the merging version ($v'$).
\end{proof}

>>>>

Secondly, we impose certain pre-conditions on the merging branches to
preempt the structure shown in Fig.~\ref{fig:external-lcas}. The
intuition is as follows: consider the branch $b_1$ at the instance of
merging $b_3$. Since it has already merged $b_2$, a version on $b_2$
($v_{22}$) could be a common ancestor for a version on $b_1$
($v_{12}$) and some other version (call it $v$). Now, if $b_1$ merges
$b_3$, same could be true of $b_3$ and $b_1$: a version on $b_3$
($v_{32}$) could be a common ancestor for the new version on $b_1$
($v_{13}$), and the other version $v$. Since $v_{22}$ is also an
ancestor of $v_{13}$, and both ancestors are not ordered by the
ancestor relation, $v_{13}$ and $v$ end up with two LCAs. In
Fig.~\ref{fig:external-lcas}, the role of $v$ is played by the version
$v_{44}$. We observe that this scenario can be prevented if, when
merging $b_3$, $b_1$ insists on an ancestor relation between the
merged version ($v_{22}$) of the $b_2$, the previously merged branch,
and the latest version of $b_3$, the current merging branch. We call
the last merged version the \emph{external locus} of the branch. By
requiring that, for every branch $b$, the external locii of $b$ at
various points in time (i.e., external locus of every prefix of $b$)
be totally ordered, we effectively enforce the invariant that for any
two common ancestors $v_1$ and $v_2$ between two versions, there
exists another common ancestor $v_3$ that succeeds $v_1$ and $v_3$ in
the ancestor relation, thereby preventing the possibility of multiple
LCAs.

%% <false>
%% Multiple versions that merge any pair of versions are totally
%% ordered. That is, if $v_1$ and $v_2$ are ancestors of $v_3$ and
%% $v_4$, then $v_3$ and $v_4$ are ordered by the ancestor relation.
%% </false>

%% If $v_1$ and $v_2$ are ancestors of $v_3$ and $v_4$, then there
%% exists a $v_5$, a successor of $v_1$ and $v_2$, and an ancestor of
%% $v_3$ and $v_4$
%% If I follow one of your locii, or you follow one of my locii, then
%% we both have same set of external common ancestors.

Let us see how the aforementioned restriction allows us to merge the
branches in Fig.~\ref{fig:external-lcas} without ever creating
multiple LCAs. The legal branching history is shown in
Fig.~\ref{fig:legal-merge}. First, like in
Fig.~\ref{fig:external-lcas}, $b_2$ is merged into $b_1$ to create
$v_{12}$, and $b_3$ is merged into $b_4$ to create $v_{42}$. Thus, the
external locus of $b_1$ is $v_{22}$, and that of $b_4$ is $v_{32}$.
Observe that now $b_1$ cannot merge $b_3$, neither can $b_4$ merge
$b_2$ because $v_{22}$ and $v_{32}$ are not related by the ancestor
relation. Same applies for $b_1$ and $b_4$ because neither one's
latest version follows from other's external locus.  However, $b_2$
and $b_3$ can merge among themselves. Let us say they indeed merge to
to create a version $v_{23}$ on $b_2$. The branch $b_2$ is now
eligible to be merged into $b_1$ and also $b_4$. These merges do occur
in Fig.~\ref{fig:legal-merge} to create versions $v_{13}$ and
$v_{43}$, and making $v_{23}$ the external locus for both $b_1$ and
$b_4$. Thus, $b_1$ can now merge into $b_4$ to create version $v_{44}$
on $b_4$.

We now formalize the intuitions described above to precisely define
the notion of mergeability:

@@

\begin{definition} [\bfseries External Locus]
Given a branch $b$ and a version $v\in b$, external locus ($v_o$) of
$v$ is an external ancestor that is not an ancestor of any other
external ancestor of $v$. That is, $\under{H}{v_o \preceq v}$, and
there does not exist a $v_o' \not\in b$ such that $\under{H}{v_o'
\preceq v}$ and $\under{H}{v_o \preceq v_o'}$. 
\end{definition}

@@

Rule \rulelabel{E-Pull-Wait} of the reduction relation
(Fig.~\ref{fig:opsem}) is enabled only if the latest version ($v'$)
of thread $t'$'s branch is mergeable into the latest version ($v$) of
thread $t$'s branch (i.e., $\under{H}{v' \mbleto v}$). Merge computes
the latest common ancestor of $v$ and $v'$, which is 
guaranteed to be unique as per the following theorem\footnote{Proof
included in the appendix}:

\begin{theorem} [\bfseries Unique LCA]
Every pair of versions $v_1$ and $v_2$ in a legal branch history $H$ 
have a unique least common ancestor. 
\end{theorem}

The \rulelabel{E-Pull-Wait} uses the function \C{lca} to compute the
LCA of (the latest versions on) a pair of branches. The definition of
the function is standard, hence not discussed.
